---
output: pdf_document
---

# Self Organizing Maps (SOM): Example using RNAseq reads
## Part 3a: Visualizing superSOMs

Getting the SOMs is easy, the hard part is interpreting. You can approach these types of analyses in many ways, but below if how I approached understanding my work. I wrote a bunch of functions, that likely only work with my data, but you can go into `sSOM_functions.R` to see how these functions are made to modify for your data.

## Required Libraries and source code

```{r}
library(VennDiagram)
library(ggplot2)
library(reshape)
library(kohonen)
library(goseq)
library(GO.db)
library(knitr)

source("sSOM_functions.R")
```


## Read in data from previous analysis. 

```{r}
plot.data <- read.table("../data/ssom.data.txt",header=TRUE)
names(plot.data)
head(plot.data)
```


## Visualization

You can examine each cluster with these functions, just change the number in the function.

```{r}
clusterVis(1)
clusterVis_PCA(1)
clusterVis_PCAsub(1)
clusterVis_line(1)
y <- genesInClust(1, plot.data, annotation)
kable(y, format = "latex", booktabs = TRUE)
clusterGO(1)
```




